---
title: "Final Project"
author: "Catherine Chalikian"
date: "4/21/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r}
library(tidyverse)
library(ggdag)
library(broom)
library(causaldata)
library(rsample)
library(visdat)
library(mice)
library(gtsummary)
library(tipr)
library(glue)
library(survey)
library(smd)
```



```{r}
dat = read.csv("FitbitsAndGradesData_Fall2017.csv")
names(dat) = tolower(names(dat))
head(dat)
```
## Do regular steps of > 10,000 average steps per day cause higher GPA in 2017 Freshman at Oral Roberts University?

### Some particulars

To answer this question, we will look at a dataset of 581 incoming Freshman at Oral Robert University (ORU) in 2017. The dataset comes from a class called “Introduction to Whole Person Education”, which required several health measures, including a timed mile, daily steps tracked by Fitbit, and daily active minutes in different heart rate zones. The 10,000 step mark is both from common recommendations, as well as the course recommendation. To ensure that this was a reasonable measure--that there would be sufficient treatment and control populations--we looked at the histogram below. There are indeed numnerous participants in both groups.

```{r, results = FALSE, warning=FALSE}
ggplot(dat)+
  geom_histogram(bins = 30,
                 aes(x = steps),
                 fill = "turquoise3")+
  geom_label(
    label = "Average Daily Steps",
    x = 16500,
    y = 50
  )+
  labs(title = "Frequency of Average Daily Steps")
```

## Missing Data? Nope.
```{r}
# visualize data
dat %>% vis_dat()
```

It appears that we no missing data, so we do not need to impute it. Let's move on and brainstorm how these variables might fit together. However, we do need to create a variable denoting treatment or control. As of now, our exposure *steps* is a continuous variable, but we want to know whether or not the participant is in the treatment group or not.

```{r}
# make our exposure var stepsT
dat$stepsT = ifelse(dat$steps > 10000, 1, 0)

# make gender factor var
dat$gender = factor(dat$gender)
head(dat)
```


## What are the relationships of the variables?

Our exposure, the average number of daily steps, could depend on a number of things, including activity level. The variables "fatburn," "cardio," and "peak" measure the number of minutes in the following respective heart rates when compared to max heart rate: 50-69%, 70-84%, and >85%. This means that each of the variables is accounting for different activity, so we will include all of them. Additionally, stress levels ("stress") could impact how many steps a student takes. There is a small distribution of Age, given that this dataset is about Freshman specifically, but it may make a difference. Gender may play a role. And finally distance from campus would likely impact number of steps. The ORU course also calculated a single score called life.ccore based on a 35-question survey. This included more inclusive elements of health, such as alcohol and drug use, psychological condition, spritiruality, and nutrition in addition to fitness. The lower the score, the healthier the person.

Then, our outcome GPA could be affected by age, gender, study habits ("study"), major, and life.score.

Based on gender stereotypes, we think that cardio, fatburn, and peak could be influenced by gender.

```{r, results = FALSE, warning=FALSE}
ggplot(dat)+
  geom_histogram(bins = 30,
                 aes(x = age),
                 fill = "turquoise3")+
  labs(title = "Frequency of Freshman Ages")
```
```{r}
dag = dagify(
  stepsT ~ stress + distance + peak + fatburn + cardio + minutes +
    age + gender + life.score,
  gpa ~ study + major + age + gender + life.score,
  cardio ~ gender,
  fatburn ~ gender,
  peak ~ gender,
  exposure = "stepsT",
  outcome = "gpa"
)

dag %>%
  ggdag()+
  geom_dag_node(color = "turquoise3")+
  geom_dag_text(color = "gray40") +
  theme_dag_gray()
```

We can visualize the previously described relashionships with the above Directed Acylcic Graph (DAG). Operations on this figure can help us identify which variables are confounders, influencing both the exposure and the outcome, as well as identify which covariates we need to adjust for to effectively close off this influence. So, that's what we'll do next with our open paths and adjustment sets.

```{r}
dag %>% ggdag_paths()+
  geom_dag_node(color = "turquoise3")+
  geom_dag_text(color = "gray40") # add labels
```

There are numerous open or backdoor paths that could lead us to a spurious result. Now let's see what we should adjust for to avoid such calamity.

```{r}
dag %>% ggdag_adjustment_set()+
  geom_dag_text(color = "gray40") # add labels
  
```

Our adjustment set is composed of the covariates *age*, *gender*, and *life.score*. We know that if assignment to treatment is unconfounded given the full set of covariates, then the assignment is also unconfounded by conditioning only on the propensity score, so we condition on the propensity score. This means that it's time to build a propensity score model!

### Propensity Scores and More

```{r}
# propensity score, unweighted
glm1 = glm(stepsT ~ age + gender + life.score,
           data = dat,
           family = binomial())

# adding propensity scores to model
dat = glm1 %>% augment(type.predict = "response",
                       data = dat)
```

Let's see how our populations compare without weighting.

```{r}
svy_des = svydesign(
  ids = ~1,
  data = dat
)

table1 = svy_des %>% tbl_svysummary(
  by = stepsT,
  include = c("age", "gender", "life.score")
) %>%
  add_overall()
table1
```

The treatment population is larger than the control. Since the data is on an incoming Freshman class, it is unsurprising that the distributions of age are the same. With males as "0", we see that more females were assigned to the treatment group.

We are going to use Average Treatment Effect of the Control (AT) to weight these measures, because we want to compare the causal effect on GPAs between the two groups.

```{r}
# add at3 weights
dat = dat %>%
  mutate(w_ate = stepsT/.fitted + 
           (1 - stepsT)/(1 - .fitted))
```

```{r, results = FALSE, warning=FALSE, message = FALSE}
svy_des = svydesign(
  ids = ~1,
  data = dat,
  weights = ~w_ate
)

wtable1 = svy_des %>% 
  tbl_svysummary(by = stepsT,
                 include = c("age", "gender", "life.score")) %>%
  add_overall()

wtable1
```

Let's see how the populations look now.

```{r, results = FALSE, warning=FALSE}
# prepare data for unweighted mirror plot
df_plot_ps = dat %>%
  tidyr::pivot_wider(names_from = stepsT,
                     values_from = .fitted,
                     names_prefix = "steps_p")
# plot
ggplot(df_plot_ps)+
  geom_histogram(bins = 25,
                 aes(x = steps_p1),
                 fill = "turquoise3")+
  geom_histogram(bins = 25,
                 alpha = .5,
                 aes(x = steps_p1,
                     weight = dat$w_ate),
                 fill = "turquoise3")+
  geom_histogram(bins = 25,
                 aes(x = steps_p0,
                     y = -stat(count)),
                 fill = "tomato3")+
  geom_histogram(bins = 25,
                 alpha = .5,
                 aes(x = steps_p0,
                     weight = dat$w_ate,
                     y = -stat(count)),
                 fill = "tomato3")+
  scale_y_continuous("count", label = abs) + 
  scale_x_continuous("propensity score") +
  geom_label(
    label = "> 10,000 Steps", 
    # x and y gives where the label will land
    x = 0.4,
    y = 35,
  ) + 
  geom_label(
    label = "< 10,000 Steps",
    x = 0.4, 
    y = -42
  ) + 
  labs(title = "Weighted Distributions of High and Low Volume Steppers")

```

Who is that High Step person with the lowest propensity score? I suspect this has to do with age, but we'll check it out.

```{r}
dat %>% filter(.fitted < .3)
```


### Checking out the model

Okay, we weighted the populations so that each look like the treated group, but is it a valid model? We're going to run some tests to find out.

#### Love Plot

```{r,results = FALSE, warning=FALSE}
# add smds to model
smds = dat %>%
  summarize(
    across(
      c("age", "gender", "life.score"),
      list(unweighted = ~smd(.x, stepsT)$estimate,
           att.weighted = ~smd(.x, stepsT, w_ate)$estimate)
    )
  )

plot_df_smds = smds %>%
  pivot_longer(
    everything(),
    values_to = "SMD",
    names_to = c("variable", "Method"),
    names_pattern = "(.*)_(.*)") %>%
  arrange(desc(Method), abs(SMD)) %>%
  mutate(variable = fct_inorder(variable))

ggplot(
  data = plot_df_smds,
  aes(x = abs(SMD), y = variable,
      group = Method, color = Method)
) +
  geom_line(orientation = "y")+
  geom_point()+
  geom_vline(xintercept = .1,
             color = "black", size = .1)+
  ggtitle("Love Plot ATE vs Unweighted")
```

It looks like the weighting brought the covariates closer to balance. All covariates are far below the rule of thumb 0.1 mark.

#### ECDF

```{r,results = FALSE, warning=FALSE}
# create weighted cdf plot for age and income

# df for exposed, age
ecdf_1a = dat %>%
  filter(stepsT == 1) %>%
  arrange(age) %>%
  mutate(cum_pct = cumsum(w_ate)/sum(w_ate))

# df for unexposed, age
ecdf_0a = dat %>%
  filter(stepsT == 0) %>%
  arrange(age) %>%
  mutate(cum_pct = cumsum(w_ate)/sum(w_ate))

#plot
ggplot(ecdf_1a, aes(x = age, y = cum_pct)) +
  geom_line(color = "Turquoise3") +
  scale_color_manual("> 10,000 Steps",
                     values = c("Blue", "Brown"),
                     labels = c("No", "Yes"))+
  geom_line(data = ecdf_0a,
            aes(x = age, y = cum_pct),
            color = "Blue") +
  xlab("Age") +
  ylab("Proportion <= x")+
  ggtitle("Weighted ECDF for Age")
```

Our continuous varaible *age* matches well, both before and after weighting. There doesn't seem to be any non-linearity present, so we will proceed to estimation!

### Cause Estimates

```{r}
dat %>%
  summarise(
    ate = sum(gpa*stepsT*w_ate)/sum(stepsT*w_ate) - sum(gpa*(1 - stepsT)*w_ate)/
      sum((1-stepsT)*w_ate)
  )
```

This is our point estimate. On average, a ORU freshman who had daily average steps over 10,000 had a GPA 0.5 higher. That sounds impressive, but how certain are we about this?

```{r}
fit_ipw <- function(split) {
  .df <- analysis(split)
  
  # fit propensity score model
  propensity_model <- glm(
    stepsT ~ age + gender + life.score,
    family = binomial(), 
    data = .df
  )
  # calculate inverse probability weights
  .df <- propensity_model %>% 
    
    # get propensity scores 
    augment(type.predict = "response", data = .df) %>% 
    # add ATE to df
    mutate(wts = stepsT / .fitted + (1 - stepsT) / (1  - .fitted)) 
  
  # fit outcome model
  lm(gpa ~ stepsT, data = .df, weights = wts) %>%
    tidy()
}

#fit ipw model
ipw_results = bootstraps(dat, 1000, apparent = TRUE) %>%
  mutate(results = map(splits, fit_ipw))

# get t-statistic-based CIs
boot_estimate = int_t(ipw_results, results) %>%
  filter(term == "stepsT")

boot_estimate
```

We are 95% confident that the true causal effect of a daily average of more than 10,000 steps for ORU freshman is (0.3749, 0.6204). There is strong evidence that high volume daily steps has a positive impact on a student's GPA.


### But...What if...?

It seems that, conditioning on *age*, *gender*, and *life score*, we have a causal effect, but what if we missed a confounder? How would the model change? It's time for sensitivity analysis. We are going to look at different standardized mean differences (SMDs). 

```{r,results = FALSE, warning=FALSE}
smds = c(.25, .5, .75)
tips1 = tip_lm(effect = .5,
        smd = smds) 
tips1
```

```{r}
tips1 %>%
  ggplot()+
  geom_line(aes(x = smds, y = outcome_association))+
  xlab("SMDs") +
  ylab("Outcome Association")+
  ggtitle("Outcome Association vs. SMD")
```


The above analysis uses the point estimate of 0.5 and confidence of 95%. When the SMD is smaller, measured at 0.25, if we had one missing (unmeasured) confounder, the confounder would need to have an outcome association of 2 to "tip" the outcome effect to 0. At this point, the observed effect would no longer be significant. If the missing confounder had an SMD of 0.5, it would need an association of 1 to render the outcome effect null. Along the same lines, with a large SMD of 0.75, the unmeasured confounder would need an outcome association of 0.667 to negate the outcome effect.

Because we are looking at a causal effect on GPA, there is a relatively small domain. Typically, GPAs range from 0 to 4.0, so an outcome association of 2 and 1 is large. We would need a huge, unmeasured confounder to nullify our results.

### Next Steps

It's pretty cool that we found a causal effect, but I wonder, does a large daily step volume cause some phsyiological change that improves cognition and therefore GPA? Or is is it specifically this level of daily steps that impacts GPA? I other words, what is the causal mechanism? We would need to do mediation analysis to determine this. 





